---
title: "Describe and Visualize Data"
---

## Introduction

An important part of data analysis is to describe and visualize the data before building models. We call it **Exploratory Data Analysis (EDA)**.

In this chapter, we separate EDA into two parts:

1.  **Univariate analysis** — examining one variable at a time.
2.  **Bivariate analysis** — examining the relationship between two variables.

Open the **`5_EDA.qmd`** file located in the **`Quarto_files`** folder. As you work through this chapter in the course compendium, copy the code examples into your Quarto document and run them yourself. Try to understand **what each line of code does, what changes it makes to the data, and why we might want to make that change**. You will use many of these techniques again later in the course.

We start by clearing the Environment, loading the required packages, loading the course-specific EDA functions, and importing the cleaned dataset.

```{r}
#| message: false
#| echo: false

# Clear environment
rm(list = ls())

# Load packages and functions
library(tidyverse)
library(scales)

# The file `EDA_functions.R` contains helper functions that we will use to create consistent plots throughout the course.
source("Functions/EDA_functions.R")

# Import the cleaned dataset
df_clean <- read.csv("Data/df_clean.csv")

```

Take a first look at the data:

```{r}
head(df_clean)
```

You can inspect the structure and produce a general summary:

```{r}
summary(df_clean)
```

## Univariate Descriptives

Univariate analysis means analyzing **one variable at a time**. Its important to remember that the type of variable we are working with determines which summary statistics, tables, and plots are appropriate.

For a numerical variable, we are usually interested in:

-   center,
-   spread,
-   shape,
-   and unusual observations.

For a categorical variable, we are usually interested in:

-   categories,
-   frequencies,
-   and proportions.

Lets go through some code that will do this for us.

### Numerical variables

We begin with numerical variables such as `age` and `income`.

We can summarize and describe the main characteristics of numerical variables. The following R functions allow us to calculate common measures of center, spread, and range.

```{r}
# Minimum value: min() returns the smallest value.
min(df_clean$age)

# Maximum value: max() returns the largest value.
max(df_clean$age)

# Range (minimum and maximum): range() returns both the minimum and maximum values.
range(df_clean$age)

# Mean: mean() calculates the arithmetic mean (average).
mean(df_clean$age)

# Median: median() returns the middle value when the observations are ordered.
median(df_clean$age)

# Standard deviation: sd() calculates the standard deviation, which describes how spread out the observations are around the mean.
sd(df_clean$age)

# Variance: var() calculates the variance, another measure of spread.
var(df_clean$age)

# Quantiles: quantile() returns selected percentiles of the distribution, including the first quartile (25%), median (50%), and third quartile (75%).
quantile(df_clean$age)
```

We can also use summary() to obtain several important summary statistics at once:

```{r}
summary(df_clean$age)
```

#### Missing values

If a variable contains missing values (NA), many of these functions will return NA. We can tell R to ignore missing values by adding na.rm = TRUE:

```{r}
mean(df_clean$age, na.rm = TRUE)

sd(df_clean$age, na.rm = TRUE)

range(df_clean$age, na.rm = TRUE)
```

Here, na.rm = TRUE means remove the NAs before performing the calculation.

#### Histogram

A histogram shows the distribution of a numerical variable. The function we will use in this course requires four arguments: the variable name (including the data frame), the number of bins, the label for the x-axis, and the label for the y-axis.

Let’s try an example by looking at the distribution of age:

```{r}
histogram(
  df_clean$age,
  bins = 40,
  xlab = "Age",
  ylab = "Frequency",
  suffix = ""
)
```

When interpreting a histogram, consider the following questions:

-   Where are most of the observations located?
-   Is the distribution symmetric or skewed?
-   Is there one clear peak or several peaks?
-   Are there any gaps in the distribution?
-   Are there any unusual or extreme observations?

#### Boxplot for one numerical variable

A boxplot summarizes the center, spread, and potential outliers of a numerical variable. In this course, we will use the following code to produce a simple boxplot for one variable.

```{r}
boxplot_simple(
  df_clean$income,
  ylab = "Income",
  suffix = "kr"
)
```

The line inside the box represents the median. The box contains the middle 50% of the observations, from the first quartile (Q1) to the third quartile (Q3). The lines extending from the box are called whiskers, and observations beyond the whiskers are considered potential outliers.

The individual data points are also overlaid on the boxplot, allowing us to see the actual observations behind the summary.

When interpreting a boxplot, consider the following questions:

-   Where is the median located?
-   Where are the middle 50% of the observations?
-   How spread out are the observations?
-   Does the distribution appear symmetric or skewed?
-   Are there any potential outliers beyond the whiskers?
-   If comparing groups, how do their centers, spreads, and outliers differ?

### Categorical variables

For categorical variables, we usually examine how observations are distributed across categories. We will use `group` as an example.

#### Frequency table

A frequency table shows the number of observations in each category.

```{r}
table(df_clean$region)
```

#### Frequency table with percentages

We can also calculate percentages:

```{r}
df_clean |>
  count(region) |>
  mutate(
    percent = n / sum(n) * 100
  )
```

This makes it easier to compare the relative size of the categories.

#### Bar chart

A bar chart provides a visual representation of category frequencies.

```{r}
ggplot(df_clean, aes(x = region)) +
  geom_bar() +
  labs(
    title = "Number of observations by group",
    x = "Group",
    y = "Number of observations"
  ) +
  theme_minimal()
```

When interpreting a bar chart, ask:

-   Which category is most common?
-   Which category is least common?
-   Are observations distributed evenly across categories?

## Bivariate analysis

Bivariate analysis means examining **two variables at the same time**. The appropriate method depends on the types of the two variables. Common combinations are:

-   numerical + numerical,
-   numerical + categorical,
-   categorical + categorical

### Numerical + numerical

When both variables are numerical, we are often interested in whether they are related.

#### Scatterplot

A scatterplot shows the relationship between two numerical variables. For example, age and income:

```{r}
scatterplot(
  df_clean,
  age,
  income,
  xlab = "Age",
  ylab = "Income"
)
```

When interpreting a scatterplot, ask:

-   Is there a positive or negative relationship?
-   Is the relationship approximately linear?
-   How strong does the relationship appear?
-   Are there unusual observations?
-   Are there clusters or other patterns?

#### Correlation

Correlation measures the strength and direction of a linear relationship between two numerical variables. Correlation describes an association. It does not by itself establish direction (causation).

```{r}
cor(
  df_clean$age,
  df_clean$income,
  use = "complete.obs"
)
```

The correlation coefficient ranges from `-1` to `1`.

-   Values close to `1` indicate a strong positive linear relationship.
-   Values close to `-1` indicate a strong negative linear relationship.
-   Values close to `0` indicate a weak linear relationship.

### Numerical + categorical

When one variable is numerical and the other is categorical, we often want to compare the distribution of the numerical variable across groups.

#### Grouped histogram

For example, we can compare income distributions across groups:

```{r}
histogram_df(
  df_clean,
  xvar = income,
  group_var = region,
  xlab = "Income",
  ylab = "Frequency",
  suffix = " kr",
  bins = 40
)
```

This allows us to examine whether the groups have different distributions.

#### Boxplot by group

A grouped boxplot is often easier to compare across categories.

```{r}
boxplot_grouped(
  df_clean$age,
  df_clean$region,
  xlab = "Region",
  ylab = "Age",
  suffix = ""
)
```

When interpreting the grouped boxplot, compare:

-   medians,
-   spread,
-   potential outliers,
-   and differences between groups.

### Categorical + categorical

#### Cross Table

When both variables are categorical, we can use a **cross table**.

```{r}
table(df_clean$age_cat, df_clean$region)
```

A more readable version can be created with:

```{r}
df_clean |>
  count(age_cat, region)
```

To examine percentages within each age category:

```{r}
df_clean |>
  count(age_cat, region) |>
  group_by(age_cat) |>
  mutate(
    percent = n / sum(n) * 100
  )
```

When interpreting a cross table, ask whether the distribution across one categorical variable appears to differ across the categories of the other variable.

## Choosing the appropriate method

A useful rule is to first identify the **type of variable or variables** you are working with.

| Analysis | Variable type | Useful methods |
|:---|:---|:---|
| Univariate | Numerical | Summary statistics, histogram, density plot, boxplot |
| Univariate | Categorical | Frequency table, percentages, bar chart |
| Bivariate | Numerical + numerical | Scatterplot, correlation |
| Bivariate | Numerical + categorical | Grouped histogram, grouped boxplot |
| Bivariate | Categorical + categorical | Cross table, percentages |

## A simple EDA workflow

A useful workflow is:

``` text
Identify the question
        ↓
Identify the variable type(s)
        ↓
Calculate appropriate summaries
        ↓
Create an appropriate visualization
        ↓
Describe the pattern
        ↓
Look for unusual observations
        ↓
Decide what to investigate next
```

EDA is not only about producing plots and tables. The goal is to use them to **understand the data and identify patterns that matter for the business or research question**.

## Key takeaway

The key distinction is:

> **Univariate analysis = one variable at a time.**

> **Bivariate analysis = two variables and the relationship between them.**

Choose the summary statistic, table, or plot based on the type of variable you are analyzing.

The goal is not simply to create a figure or table. The goal is to use the figure or table to **describe the data, identify patterns, and support further analysis**.

------------------------------------------------------------------------
