---
title: "Linear Regression: Lab 1"
subtitle: "Introduction to Supervised Learning and Prediction"
format: html
execute:
  warning: false
  message: false
---

## Introduction

In the previous labs, we worked with k-means clustering. K-means is an unsupervised learning method. There is no outcome variable that the model is trying to predict. Instead, the algorithm searches for patterns in the data and groups similar observations together.

We now move to **supervised learning**. In supervised learning, we have an **outcome variable** that we want to explain or predict using one or more **predictor variables**.

In this lab, we will introduce a basic modeling workflow for supervised learning. This workflow differs somewhat from the workflow we used in the previous lab, where we worked with unsupervised learning. We will work with the following overarching steps:

1.  identify the outcome and predictor variables
2.  split the data into training and test data
3.  fit different models using the training data
4.  compare how well the models fit the training data
5.  select a model
6.  use the selected model to make predictions on the test data
7.  evaluate the model's predictive performance using the test data

We will begin this workflow using **linear regression**. In chapter 12 and 13 we will use a similar workflow but with logistic regression.

Read Sections 3.2 and 3.3 in *An Introduction to Statistical Learning - with Applications in R* before working with the lab. Also read relevant parts of Chapter 6 in *An Introduction to Statistical Learning - with Applications in R* before working with the lab.

### Learning objectives

After completing this lab, you should be able to:

- distinguish between outcome and predictor variables
- explain the basic idea behind linear regression
- fit a simple linear regression model in R
- interpret the intercept and slope
- calculate predicted values
- understand the difference between observed and predicted values
- calculate prediction errors
- explain why we separate training, and test data


## Exercise 1: Simple Linear Regression

### Prepare the R session

We begin by clearing the Environment and loading the packages used in the lab.

```{r}
# Clear Environment
rm(list = ls())

# Load packages
library(tidyverse)
library(ggplot2)

source("Functions/EDA_functions.R")
source("Functions/linear_regression_functions.R")
```

### Import the data

In the first part of this lab, imagine that we work for a company that wants to understand and predict **sales**. The company has collected information about its marketing activities. Lets import some data.

```{r}
# Import the data
df <- read.csv("Data/regression_data.csv")
```

Take a first look at the data.

```{r}
head(df)

```

Inspect the descriptive statistics.

```{r}
summary(df)

```

#### Questions

1.  How many observations are in the dataset?
2.  Which variables are included?
3.  Which variable do you think we should try to predict?
4.  Which variable could be used to predict it?
5.  Do you find anything in the data that might be problematic?

### Outcome and predictor variables

In supervised learning, we distinguish between outcome variables and predictor variables.

- The **outcome variable** is the variable we want to explain or predict.
- The **predictor variables** contain information that we use to make that prediction.

We will start with a simple linear regression model with only one predictor. Suppose we want to predict `sales` using `marketing`. In this case:

- `sales` is the outcome variable;
- `marketing` is the predictor variable.

We can write this relationship as:

$$Sales = f(Marketing)$$

This simply means that we are interested in whether information about marketing can help us understand or predict sales.

### Explore the relationship

Before fitting a model, we should always explore the data. Because both variables are quantitative, a scatter plot is useful.

```{r}
scatterplot(
  df,
  marketing,
  sales,
  xlab = "Marketing",
  ylab = "Sales"
)


```

#### Questions

1.  Does there appear to be a relationship between marketing and sales?
2.  Is the relationship positive or negative?
3.  Does the relationship appear approximately linear?

### The linear regression model

Linear regression describes the relationship between an outcome variable and one or more predictor variables using a line. For a simple linear regression with one predictor, the fitted model can be written as:

$$\hat{y} = b_0 + b_1x$$

where:

- $\hat{y}$ is the predicted value of the outcome
- $b_0$ is the **intercept**
- $b_1$ is the **slope**
- $x$ is the predictor variable

For our example:

$$\widehat{Sales} = b_0 + b_1 Marketing$$

The model tries to find the line that best represents the relationship between marketing and sales. The position and slope of this line are determined by two coefficients, $b_0$ and $b_1$, which are estimated from the data. Together, these coefficients define the best-fitting line—the line that minimizes the squared differences between the observed sales values and the values predicted by the model.

### Fit the model

We use the `lm()` function to estimate a linear regression model.

```{r}
model <- lm(
  sales ~ marketing,
  data = df
)

```

View the estimated coefficients.

```{r}
coef(model)
```

You can also inspect a more detailed model summary.

```{r}
summary(model)
```

#### Questions

1.  What is the estimated intercept?
2.  What is the estimated slope for `marketing`?
3.  Is the slope positive or negative?
4.  What does the slope tell us about the relationship between marketing and sales?

### Interpret the model

Suppose the estimated model is:

$$\widehat{Sales} = 20 + 2.5 \times Marketing$$

The intercept of 20 means that the model predicts sales of 20 when marketing is 0. The slope of 2.5 means that a one-unit increase in marketing is associated with an increase of 2.5 units in predicted sales.

#### Question

1.  Using the coefficients from your own model, write the estimated regression equation.

### Add the regression line

We can add the fitted regression line to the scatter plot.

```{r}
plot_regression_line(
  data = df,
  model = model,
  xvar = marketing,
  yvar = sales,
  xlab = "Marketing",
  ylab = "Sales"
)
```

#### Questions

1.  Does the regression line appear to represent the relationship reasonably well?
2.  Are all observations located exactly on the line?
3.  Why do you think observations differ from the regression line?

### Make predictions

Once we have fitted a model, we can use it to make predictions. Suppose marketing is equal to 30.

```{r}
new_customer <- data.frame(
  marketing = 30
)

predict(
  model,
  newdata = new_customer
)
```

#### Questions

1.  What sales value does the model predict when marketing is: 30, 35, 55?
2.  Use your regression equation to calculate the same prediction by hand.

### Predicted values for all observations

We can also calculate a prediction for every observation in the dataset.

```{r}

df <- df |>
  mutate(
    predicted_sales = predict(model)
  )
```

Inspect the results.

```{r}
head(df)
```

We now have:

- the observed value: `sales`;
- the model's prediction: `predicted_sales`.

### Prediction error

A model will normally not predict every observation perfectly. The difference between the observed value and the predicted value is called a **residual** or prediction error.

$$Error = Observed - Predicted$$

Calculate the errors.

```{r}
df <- df |>
  mutate(
    error = sales - predicted_sales
  )
```

Inspect the results.

```{r}
head(df)

```

We can plot errors with a residual plot:

```{r}
# Plot the prediction errors
plot_residuals(
  data = df,
  xvar = marketing,
  errorvar = error,
  xlab = "Marketing"
)
```

Each point shows the prediction error for one observation. The dashed horizontal line represents an error of zero, meaning that the prediction was exactly correct.

- Points above zero are observations where the model predicted too little.
- Points below zero are observations where the model predicted too much.
- Points close to zero represent relatively accurate predictions.

#### Questions

- Are the errors generally centered around zero?
- Are there observations with particularly large errors?
- Do you see any systematic pattern in the errors?
- Find an observation with a positive error. What does a positive error mean?
- Find an observation with a negative error. What does a negative error mean?
- Would a perfect model have large or small errors?

### Measure prediction error

To evaluate a model, we need to summarize its prediction errors. A common measures is **Mean Squared Error (MSE)**. MAE calculates the average squared difference between observed and predicted values.

For the MSE, **lower values** indicate better predictive performance. Ideally, we want it to be as small as possible. A value of zero would mean that the model predicts every observation perfectly. We will return to evaluating models using the MSE in the next lab.

Try calculate the MSE below (remember that we already calculated the prediction errors).

```{r}

mse <- mean(df$error^2)

mse
```

#### Questions

1.  What is the MSE of the model?
2.  Would you prefer a model with a smaller or larger prediction error?


## Exercise 2: Multiple linear regression

Usually, we want to use much more information to predict the outcome than what can be captured in a single variable. Therefore, we use multiple linear regression, where we have several predictor variables.

For this part, we will work with a real dataset. The dataset, containing the prices and other attributes of almost 54,000 diamonds, is included in the `ggplot2` package. A description of the data can be inspected using the command `?diamonds`.

```{r}
# Import the data
df2 <- diamonds

# Get variable descriptions
?diamonds
```

#### Questions

1.  How many variables are there? Which ones are quantitative (numeric) and which ones are qualitative (categorical)?
2.  Which variables do you think would be most useful for predicting the `price` of diamonds?

### Fitting the first model

Normally, we would start by exploring the relationship between the outcome, `price`, and the potential predictors in the dataset (see the steps in the first half of the lab). Even though they are important, we will skip these steps for this part of the lab so that we have time to build more complicated models. You should try to do those steps on your own after the lab, using the code presented in the first half.

For this modelling process, we will focus more on predicting values of the outcome and less on explaining the relationship between the predictors and the outcome. We start by fitting a simple linear regression model with `carat` as the predictor:

```{r}
model1 <- lm(
  price ~ carat,
  data = df2
)


```

We can quickly inspect the model using the `summary()` function, or just look at certain aspects of the summary. We will look at the adjusted coefficient of determination, $R^2$(adjusted). This is a measure of how well the model explains the variation in the outcome, but adjusted for the number of terms (predictors) in the model. A higher adjusted $R^2$ means that a model is better at explaining variation in the response.

```{r}
adj_R2_model1 <- summary(model1)$adj.r.squared
adj_R2_model1
```

The adjusted $R^2$ can only take values between 0 and 1, so a value of 0.85 is pretty high. As before, we can also calculate the prediction errors for this model. This time, we will **not** save the values in our data frame, but as separate objects instead:

```{r}
predictions_model1 <- predict(model1)

error_model1 <- df2$price - predictions_model1
```

Next, calculate the MSE for the first model:

```{r}
mse_model1 <- mean(error_model1^2)

mse_model1

```

For the MSE, a smaller value is better because it means the predictions are more correct, on average. But without having anything to compare it too, we can't really say if the value we got is "good" or "bad". Note that it wouldn't make sense to compare the MSE we got here with the one from exercise 1, because the two models are predicting completely **different** outcomes with **different** data sets. What we can compare, however, is how good different models based on the **same** data are at predicting the **same** outcome, i.e., diamond prices.

### Fitting several models

So, let's build a bigger, more complicated model for diamond prices, using all the predictors available. If we want to use all variables in a dataframe as predictors, we can use the formula `y ~ .` when specifying the regression model. This way, R will automatically add all variables - except the response, of course - as predictors in the model. (This is the reason we didn't save the predictions from the previous model in the data frame).

```{r}
model2 <- lm(
  price ~ .,
  data = df2
)
```

As for the first model, we aren't really interested in interpreting the coefficients here, but only in how well our model performs. Calculate the adjusted $R^2$, prediction errors and MSE for this second model as well, and save the values as new objects using the same naming scheme. We will compare those values later.

```{r}
adj_R2_model2 <- summary(model2)$adj.r.squared

predictions_model2 <- predict(model2)

error_model2 <- df2$price - predictions_model2

mse_model2 <- mean(error_model2^2)
```

Lastly, we will build an even more complicated model. This third model will have an interaction term between `carat` and `cut`. Specifying a model like this means that we think that the relationship between carat and price is different for diamonds with different values of the variable cut. For example, it could be that the price of a diamond with an "ideal" cut increases more with carat than it does for diamonds with a "fair" cut. You do not need to understand the details of how interaction terms work for this lab.

```{r}
model3 <- lm(
  price ~ . + carat*cut, 
  data = df2
)
```

Calculate and save the adjusted $R^2$, prediction errors and MSE for the third model.

```{r}
adj_R2_model3 <- summary(model3)$adj.r.squared

predictions_model3 <- predict(model3)

error_model3 <- df2$price - predictions_model3

mse_model3 <- mean(error_model3^2)
```

### Comparing the output

Lastly, lets create a table to compare the two values (adjusted $R^2$ and MSE) over the three different models:

```{r}
comparison <- data.frame(
  Model = c(1, 2, 3),
  adj.R2 = c(
    adj_R2_model1, 
    adj_R2_model2,
    adj_R2_model3
  ), 
  MSE = c(
    mse_model1, 
    mse_model2,
    mse_model3
  )
)

comparison
```

#### Questions

1.  Which of the three models is best at describing the variation in the outcome in our data? Which one is best at predicting the outcomes in our data?
2.  Do you see any potential problems with the modelling approach we took in the second half?

### Summary

K-means and linear regression represent two different types of machine learning.

**K-means clustering** is unsupervised:

- there is no outcome variable;
- the algorithm searches for groups in the data.

**Linear regression** is supervised:

- there is an outcome variable;
- predictors are used to explain or predict the outcome.

In this lab, we introduced part of a basic supervised learning workflow:

1.  identify the outcome and predictors;
2.  explore their relationship;
3.  fit the model
4.  examine the model's performance
5.  use the fitted model to predict the outcome

In later labs, we will fit increasingly sophisticated predictive models and examine an important challenge in supervised learning: a model that fits the training data better does not necessarily predict new observations better. This will lead us to the concepts of **overfitting, model complexity, and model selection**.
