---
title: "Lab7"
---

## Introduction

In the previous lab, we introduced **supervised learning** and the basic predictive modeling workflow. We learned how to:

1.  identify an outcome variable;
2.  identify predictor variables;
3.  fit a linear regression model;
4.  compare the performance of different models.

The main problem with the modelling approach we took in the second half of Lab6 (apart from not inspecting the data properly before modelling) was that we used the same data to **train** the models as we used to **evaluate** their performance. This can lead to a problem called **overfitting**. A more complicated model will almost always be better at predicting the outcome for the same data set and observations it was trained on, but it might not be the best for predicting the outcome for **new** observations. In practical business applications, however, predicting new observations is the entire point of building machine learning models.

In this lab, we therefore divide the data into two parts:

- **training data** — used to fit and develop models;
- **test data** — used to evaluate how well the model predicts new, unseen observations.

::: callout-important
### Why do we need test data?

A model is fitted using the training data and will therefore tend to perform well on those observations. However, good performance on the training data does not necessarily mean that the model will perform equally well on new data.

The test data contain observations that were not used to fit the model. They therefore allow us to evaluate how well the model generalizes to new observations.
:::

In this lab, we will use:

- 70% training data;
- 30% test data.

In addition to **splitting the data,** our modeling workflow will still include **model comparison and model selection** like we did in the previous lab. The new workflow will be:

**Business problem → Split data → Fit candidate models → Compare models → Select one model → Evaluate on test data → Business decision**

### Learning objectives

After completing this lab, you should be able to:

- explain the difference between training and test data;
- fit several regression models;
- calculate and compare training errors;
- compare alternative models;
- select a model based on training performance and other relevant considerations;
- evaluate the selected model using test data;
- translate model performance into a business recommendation.

### Prepare the R session

```{r}
# Clear Environment
rm(list = ls())

# Load packages
library(tidyverse)
library(ggplot2)

source("Functions/EDA_functions.R")
source("Functions/linear_regression_functions.R")
```

### Import the Data

We continue with the `diamonds` data from the previous lab.

```{r}

```

Take a first look at the data

```{r}

```

Inspect the descriptive statistics.

```{r}

```

### Explore the Data

Before building models, we should understand the variables in our data and the relationships between them. One way to explore these relationships is to create a correlation matrix.

Correlation measures the strength and direction of the linear relationship between two variables. Correlation values range from -1 to 1:

- values close to 1 indicate a strong positive relationship;
- values close to -1 indicate a strong negative relationship;
- values close to 0 indicate a weak or no linear relationship.

A correlation matrix allows us to quickly examine these relationships across several variables at the same time.

```{r}




```

#### Questions

1.  Which variables appear to be related to price?
2.  Which relationships appear positive?
3.  Which relationships appear negative?
4.  Are there predictors that appear to contain little information about price?
5.  Do any of the predictors seem to have strong correlations with each other?

### Training and Test Data

We divide the data into **training data** and **test data**.

| Dataset       | Purpose                                             |
|---------------|-----------------------------------------------------|
| Training data | Fit, develop, and compare models                    |
| Test data     | Evaluate predictive performance on new observations |

: In this lab, we use:

- 70% training data;
- 30% test data.

::: callout-important
#### Important

The models are fitted using the **training data**.

We will also initially compare how well the different models fit the training data. However, better training performance does not necessarily mean better performance on new observations.

The **test data** allow us to examine how well the selected model generalizes to observations that were not used to fit it.
:::

### Split the Data

Use the `split_data()` function to divide the data into training and test data.

```{r}





```

Create two data frames:

```{r}




```

### Model 1: A simple model

We will start by fitting the same three models as we did in the last lab, but this time only using the training data. The first is a simple linear regression model that only uses `carat` as a predictor.

```{r}



```

We can inspect a quick summary of the model, even though we will not be focusing on interpreting the coefficients this time.

```{r}


```

#### Questions

1.  Does `carat` seem to be related to `price`? If so, describe the relationship.
2.  What is the adjusted $R^2$ for this model? Is it roughly the same as last time?

### Model 2: All predictors

Next, let's fit a model using all the available predictors, and look at a summary of this as well.

```{r}




```

#### Questions

1.  Which predictors are included in Model 2?
2.  Which coefficients are positive?
3.  Which coefficients are negative?
4.  Why might Model 2 make better predictions than Model 1?

### Model 3: A very complicated model

In the previous lab, the last model we built contained an interaction term between two of the predictors. Today, we will build a model that contains pairwise interactions between **all** of the predictors. We will treat this as a kind of "black-box" model. That is to say, we will use the output of the model to make predictions, but we don't necessarily need to know exactly how these predictions are made.

**Note** that because of its size, running the code to fit the third model will take some time.

```{r}




```

We will not inspect the coefficients for this last model.

### Make Predictions on the Training Data

We now use each model to predict the observations in the **training data**.

```{r}






```

Remember that these are the same observations that were used to fit the models. We are therefore measuring how well each model fits the **training data**.

### Calculate Training Errors

Calculate the prediction errors for each model.

```{r}




```

### Compare Models Using MSE

We compare the models using the **Mean Squared Error (MSE)**.

$$MSE =
\frac{1}{n}
\sum
(y_i - \hat{y}_i)^2.$$

A smaller MSE indicates that the predictions are, on average, closer to the observed values.

### Create a Model Comparison Table

Calculate the training MSE for each model.

```{r}
# 1) We calculate the mean and save it to a new object





# 2) we create a new data frame called model_comparison






```

#### Questions

1.  Which model has the lowest training MSE?
2.  Which model has the highest training MSE?
3.  How large are the differences?
4.  Does adding more predictors and more complicated terms reduce the training error?
5.  Does the most complicated model have the lowest training MSE?
6.  Do you think the model with the lowest training MSE will also perform best on the test data?

::: callout-important
#### Training performance and model complexity

A model with a lower training MSE fits the training observations more closely. However, this does **not necessarily mean that it will make better predictions for new observations**.

As models become more flexible, they can fit the training data increasingly well. Eventually, a model may begin to fit patterns that are specific to the training observations rather than patterns that generalize to new data. This is called **overfitting**.
:::

### Evaluate the model using the test data

So far, we have only compared how well the three models perform on the training data. A more complex model will almost always have a lower **training** MSE, but that doesn't necessarily correspond to a lower **test** MSE. Next, let's see how the models perform on the test data, which wasn't used to train the models.

```{r}
# We first create predictions, but on the test data






# We calculate the errors




# We calculate the MSE




```

Add the new test MSE as a column to our comparison table.

```{r}
# We create a new data frame comparing training and test MSE






```

#### Questions

1.  Which models have the lowest and highest test MSE?
2.  How large are the differences in test MSE between models?
3.  Which models have the largest and smallest difference between training- and test MSE?
4.  Which model do you think performs best overall?

### Other considerations when selecting a model

There are several other aspects, apart from training- and test MSE, that could be important when selecting which model to implement in practice. Such as:

- model complexity
- interpretability
- whether the improvement over simpler models is meaningful
- cost of collecting data
- computational costs.

Different aspects can be more or less important in different business contexts. For example, sometimes we are only interested in the predictions of the model, and other times we might want to be able to interpret its coefficients in order to understand more about the problem at hand.

### Plot observed and predicted prices

Choose the model you think would work best in practice, based on the considerations above. Assign your selected model to `selected_model`.

For example, if you selected Model 2:

```{r}




```

Create a dataset containing the observed values, predictions, and errors.

```{r}




```

We can visualize how closely the predicted prices correspond to the observed prices in the test data.

```{r}





```

The dashed line represents perfect predictions:

$$Observed = Predicted$$

Points close to the line represent relatively accurate predictions.

#### Questions

1.  Are most observations close to the line?
2.  Are there observations with particularly poor predictions?
3.  Does the model appear to systematically predict too high or too low?
4.  Are there patterns for when the model performs better or worse?

### Plot the Prediction Errors

```{r}

# Plot the prediction errors






```

Remember:

- an error close to **0** represents an accurate prediction;
- a **positive error** means that the model predicted too little;
- a **negative error** means that the model predicted too much.

#### Questions

1.  Are the errors generally centered around zero?
2.  Are there particularly large prediction errors?
3.  Do you see a systematic pattern?
4.  What would an ideal error plot look like?

### From Model Evaluation to Business Decision

A model having a relatively low MSE does not automatically mean that the company should implement it. Management also needs to consider whether the predictions are sufficiently accurate to be **useful for the business problem**.

For example, consider:

- How costly are prediction errors?
- Is the prediction accuracy good enough for the intended decision?
- Is the model understandable?
- Is the required data available when predictions need to be made?
- Are there observations for which the model performs particularly poorly?

This means that there are two related decisions:

**Model selection**

> Which candidate model should we use?

and

**Business decision**

> Is the selected model good enough to use?

### Final Management Recommendation

Imagine that you are presenting your analysis to company management. Prepare a short recommendation.

#### Model Selection

1.  Which models did you compare?
2.  Which model did you select?
3.  Why did you select this model?
4.  What was its training MSE?

#### Test Performance

5.  What was the test MSE?
6.  How did the test MSE compare with the training MSE?
7.  Does the model appear to generalize well to new observations?

#### Business Decision

9.  Does the model appear to provide useful predictions?
10. Are there important limitations?
11. Would you recommend that the company implement the model?

Conclude with one of the following:

**Implement the model**

or

**Do not implement the model**

Briefly justify your decision.


### Summary

In the previous lab we learned the basic idea of supervised learning. In this lab, we extended that idea to **model comparison**. We learned that different models can produce different prediction errors and that model complexity does not automatically produce better predictions for new observations.

We continued to use two different datasets:

- **training data** are used to fit, develop, and compare models;
- **test data** are used to evaluate predictive performance on new observations.

An important distinction is that a low **training error** tells us that a model fits the training observations well. It does not necessarily tell us that the model will make accurate predictions for new observations. The **test error** allows us to examine how well the model generalizes to data that were not used to fit it.

Finally, predictive modeling does not end when we calculate MSE. The final question is whether the model produces predictions that are sufficiently accurate and useful to support the company's decisions.

In the next labs, we will continue to develop this predictive modeling workflow and introduce other methods.
